---
title: "Nigerian PAYE Tax Computation with nrsr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nigerian PAYE Tax Computation with nrsr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
library(nrsr)
```

## Overview

**nrsr** computes Nigerian PAYE tax under any registered tax law version.
All tax bands, relief rules, and statutory rates live in a single **law registry**
(`R/law_registry.R`). When the government amends the tax law, you only edit
that one file — no calculation logic changes anywhere else.

---

## Registered Tax Laws

```{r laws, eval=FALSE}
list_tax_laws()
```

---

## Auto-Selecting the Right Law by Date

```{r auto, eval=FALSE}
get_applicable_law("2025-06-01")   # returns "PITA"
get_applicable_law("2026-03-12")   # returns "NTA2025"

# calc_paye() also accepts law = "auto"
t_auto <- calc_paye(1624734, law = "auto")
cat("Law selected:", t_auto$law, "\n")
```

---

## Single Employee — NTA 2025

```{r single, eval=FALSE}
t1 <- calc_paye(1624734, annual_rent = 1800000)
print(t1)
```

---

## Net Take-Home Pay

```{r net, eval=FALSE}
net <- calc_net_salary(1624734, annual_rent = 1800000)
print(net)
```

---

## Compare All Registered Laws

The comparison automatically includes every law in the registry.
Add a new amendment and it appears here with no code changes.

```{r compare, eval=FALSE}
cmp <- compare_tax_laws(1624734, annual_rent = 1800000)
print(cmp)
```

---

## Tax Band Breakdown

```{r breakdown, eval=FALSE}
tax_breakdown(1624734, annual_rent = 1800000)
```

---

## Batch Payroll

```{r batch,eval=FALSE}
employees <- data.frame(
  name          = c("Emeka Obi", "Ngozi Adeyemi", "Tunde Balogun"),
  gross_monthly = c(693228.96,   1624734,          3500000),
  annual_rent   = c(0,           1800000,           2400000)
)
payroll <- calc_paye_batch(employees)
payroll[, c("name", "gross_monthly", "monthly_tax", "effective_rate_pct", "net_monthly")]
```

---

## How to Add a Future Tax Amendment

When the government amends the tax law, **only one file changes**: `R/law_registry.R`.
Copy the template block at the bottom of the registry:

```r
NTA2028 = list(
  key            = "NTA2028",
  description    = "Nigeria Tax Amendment Act 2028",
  effective_from = as.Date("2028-01-01"),
  effective_to   = NA,

  bands = data.frame(
    band_width = c(1200000, 2800000, 9000000, 13000000, 25000000, Inf),
    rate       = c(0.00,    0.15,    0.18,    0.21,     0.23,     0.25),
    label      = c("First N1.2M @0%", "Next N2.8M @15%",
                   "Next N9M @18%",   "Next N13M @21%",
                   "Next N25M @23%",  "Above N50M @25%"),
    stringsAsFactors = FALSE
  ),

  deductions = list(
    pension_employee_rate = 0.08,
    pension_employer_rate = 0.10,
    nhf_rate              = 0.025,
    nhis_rate             = 0.05
  ),

  relief_rules = list(
    cra_fixed          = NULL,
    cra_variable_rate  = NULL,
    rent_relief_rate   = 0.20,
    rent_relief_max    = 700000,    # cap raised to N700K
    tax_free_threshold = 1200000    # threshold raised to N1.2M
  )
)
```

After adding this entry, **zero other files change**:

| What works automatically | How |
|---|---|
| `list_tax_laws()` shows it | registry scan |
| `get_applicable_law("2028-06-01")` returns `"NTA2028"` | date range logic |
| `calc_paye(..., law = "auto")` uses it after 2028-01-01 | date dispatch |
| `calc_paye(..., law = "NTA2028")` computes correctly | registry lookup |
| `compare_tax_laws(...)` includes it | iterates all keys |
| `calc_paye_batch(df, law = "NTA2028")` works | passes through |

---

## References

- Nigeria Tax Act (NTA) 2025, signed June 2025
- [PaidHR — Understanding PAYE Tax Calculation in Nigeria](https://www.paidhr.com/blog/understanding-tax-computation-in-nigeria-a-step-by-step-guide)
- [Nigeria Revenue Service](https://www.nrs.gov.ng)
