---  
title: "PAYE from First Principles: A Deep Dive with nrsR"  
output: rmarkdown::html_vignette  
vignette: >  
  %\VignetteIndexEntry{PAYE from First Principles: A Deep Dive with nrsR}  
  %\VignetteEngine{knitr::rmarkdown}  
  %\VignetteEncoding{UTF-8}  
---  
  
```{r setup, include = FALSE}  
knitr::opts_chunk$set(  
  echo = TRUE,  
  collapse = TRUE,  
  comment = "#>"  
)  
library(nrsR)  
```  
  
## 1. Introduction  
  
Every month, millions of Nigerian workers see a line on their payslip  
labelled **PAYE** — money withheld before they are paid. Employers are  
legally responsible for computing that figure correctly, and getting it  
wrong means either short-changing the employee or under-remitting to the  
**Nigeria Revenue Service (NRS)**.  
  
`nrsR` exists to make that computation **exact, reproducible, and  
future-proof**. This vignette takes you from "what is PAYE?" all the way to  
running payroll for an entire workforce — and shows why `nrsR`'s design  
means a 2028 tax amendment will cost you *one file edit*, not a rewrite.  
  
## 2. Definition  
  
**PAYE (Pay-As-You-Earn)** is a system where income tax is deducted at  
source, spread across the year rather than paid in one lump sum. In Nigeria  
the amount owed is **progressive**: income is sliced into bands, and each  
band is taxed at its own rate.  
  
Two laws matter today:  
  
```{r laws}  
list_tax_laws()  
```  
  
| Term | Meaning in `nrsR` |  
|------|-------------------|  
| **Gross** | Total monthly pay before any deduction |  
| **Relief** | Amounts subtracted before tax (pension, NHF, NHIS, rent relief, CRA) |  
| **Taxable income** | Gross annual minus total reliefs |  
| **Band** | An income slice with a fixed rate |  
| **Law registry** | The single source of truth for all bands, rates, and reliefs |  
  
## 3. Storytelling  
  
Meet **Ngozi**, a Lagos product manager earning **₦1,624,734/month** who  
pays **₦1,800,000/year** in rent. In December 2025 she is taxed under the  
old **PITA** regime. On 1 January 2026 the **Nigeria Tax Act 2025  
(NTA2025)** takes effect — the first ₦800,000 of annual income becomes  
tax-free, the old Consolidated Relief Allowance (CRA) disappears, and a new  
**rent relief** appears.  
  
Ngozi wants to know: *"Will I take home more or less next year?"* With  
`nrsR`, she doesn't guess — she computes both worlds and compares them.  
  
```{r story}  
compare_tax_laws(1624734, annual_rent = 1800000)  
```  
  
The story of `nrsR` is really the story of **change**: tax laws evolve, and  
the tool is built so that evolution is cheap.  
  
## 4. Illustration / Analogy  
  
Think of progressive tax like **filling a set of stacked buckets**.  
  
Income pours in from the top. The **first bucket** (first ₦800,000 under  
NTA2025) has a tap set to **0%** — nothing drips out as tax. Once it  
overflows, income spills into the **next bucket**, whose tap is open wider  
(15%), and so on. The higher the bucket, the wider the tap.  
  
You are never taxed at your top rate on *all* your money — only on the  
portion that reaches that bucket. `tax_breakdown()` shows exactly how much  
landed in each bucket and how much dripped out:  
  
```{r buckets}  
tax_breakdown(1624734, annual_rent = 1800000)  
```  
  
## 5. Examples  
  
### One employee, full computation  
  
```{r ex-single}  
t1 <- calc_paye(1624734, annual_rent = 1800000)  
print(t1)  
```  
  
### Take-home pay (the payslip view)  
  
```{r ex-net}  
calc_net_salary(1624734, annual_rent = 1800000)  
```  
  
### Just the reliefs  
  
```{r ex-reliefs}  
calc_reliefs(1624734, annual_rent = 1800000)  
```  
  
### Let the date pick the law  
  
```{r ex-auto}  
get_applicable_law("2025-06-01")   # PITA  
get_applicable_law("2026-03-12")   # NTA2025  
calc_paye(1624734, law = "auto")$law  
```  
  
### A whole workforce at once  
  
```{r ex-batch}  
employees <- data.frame(  
  name          = c("Emeka Obi", "Ngozi Adeyemi", "Tunde Balogun"),  
  gross_monthly = c(693228.96,   1624734,          3500000),  
  annual_rent   = c(0,           1800000,           2400000)  
)  
payroll <- calc_paye_batch(employees)  
payroll[, c("name", "gross_monthly", "monthly_tax",  
            "effective_rate_pct", "net_monthly")]  
```  
  
### Format like a payslip  
  
```{r ex-format}  
format_naira(t1$monthly_tax)  
```  
  
## 6. Insights  
  
- **Effective rate ≠ top band rate.** Ngozi may touch the 18% band, but her  
  *effective* rate (tax ÷ gross) is much lower because the lower buckets are  
  cheap or free. Read `effective_rate` from `calc_paye()`.  
- **Reliefs are leverage.** Pension (8%) and rent relief directly shrink  
  taxable income. Two people with identical gross can owe different tax.  
- **NTA2025 helps lower earners most.** The ₦800K tax-free floor removes  
  low-income workers from tax entirely — verify by running a small salary  
  through both laws.  
- **Auto-dispatch removes human error.** Passing `law = "auto"` means your  
  payroll always uses the legally-correct law for the run date.  
  
## 7. Diagram — how a computation flows  
  
```  
        gross_monthly, annual_rent, toggles, law  
                          |  
                          v  
                  +----------------+  
                  |  calc_paye()   |  
                  +----------------+  
                          |  
             1. resolve law (registry)  
                          |  
                          v  
                  +----------------+  
                  | calc_reliefs() |  pension, NHF, NHIS,  
                  +----------------+  CRA / rent relief  
                          |  
             2. taxable = gross_annual - reliefs  
                          |  
                          v  
                  +----------------+  
                  | .apply_bands() |  pour income through  
                  +----------------+  the stacked buckets  
                          |  
             3. sum per-band tax  
                          |  
                          v  
        annual_tax, monthly_tax, effective_rate  
```  
  
Every arrow reads its numbers from **one place** — `.TAX_LAW_REGISTRY` in  
`R/law_registry.R`.  
  
## 8. Visual Analogy — the registry as a "law cartridge"  
  
Picture an old games console. The **console** is your calculation engine  
(`calc_paye`, `calc_reliefs`, `.apply_bands`) — it never changes. Each tax  
law is a **cartridge** you slot in: `PITA`, `NTA2025`, and one day  
`NTA2028`. Swapping the cartridge changes the whole game (bands, rates,  
reliefs) without opening up the console.  
  
Adding a future law = **printing a new cartridge** (one list entry). The  
console instantly plays it: `list_tax_laws()`, `compare_tax_laws()`,  
`calc_paye_batch()` all recognise it automatically.  
  
## 9. Mental Models  
  
Keep three models in your head:  
  
1. **Stacked buckets** → progressive bands; only overflow reaches the next tap.  
2. **Console + cartridges** → fixed engine, swappable law definitions.  
3. **Funnel** → gross goes in the top, reliefs are subtracted, taxable  
   income drops into the buckets, tax comes out the bottom.  
  
If you can picture those three, you understand both PAYE *and* the `nrsR`  
architecture.  
  
## 10. Quick Recall / Memory Hooks  
  
- **"Buckets, not blankets."** You're taxed bucket-by-bucket, not one flat  
  blanket rate.  
- **"800 is the floor."** Under NTA2025, the first ₦800K/year is tax-free.  
- **"Reliefs shrink the taxable, not the gross."** Deductions lower what's  
  taxed, not what you earn.  
- **"One law, one file."** New amendment? Edit `R/law_registry.R` only.  
- **"Auto = accurate."** `law = "auto"` always picks the right law for the date.  
  
### Function cheat-sheet  
  
| I want... | Call |  
|-----------|------|  
| Tax for one person | `calc_paye()` |  
| Take-home pay | `calc_net_salary()` |  
| Just deductions | `calc_reliefs()` |  
| Whole payroll | `calc_paye_batch()` |  
| Compare laws | `compare_tax_laws()` |  
| Band-by-band detail | `tax_breakdown()` |  
| See all laws | `list_tax_laws()` |  
| Law for a date | `get_applicable_law()` |  
| Currency string | `format_naira()` |  
  
## References  
  
- Nigeria Tax Act (NTA) 2025  
- [Nigeria Revenue Service](https://www.nrs.gov.ng)
